Parameter Recovery Validation

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: add latent time prior to parameter recovery validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

- **Time coordinates**: Latent $t^*_j$ with priors
- **Dynamics**: Same dimensionless analytical solutions
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus latent time coordinates
- **Prior**: Log-normal hierarchical priors
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

- **Time coordinates**: Latent $t^*_j$ with priors
- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus latent time coordinates
- **Prior**: Log-normal hierarchical priors
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

*To be implemented: Same dimensionless approach but with latent time coordinates that must be inferred along with other parameters.*

>>>>

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Same dimensionless analytical solutions
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

This validation study extends the dimensionless approach to include **latent time coordinates** that must be inferred along with other parameters. We implement a hierarchical time structure that maintains biological interpretability while ensuring statistical identifiability.

The model uses the same dimensionless dynamics and observation model as Validation Study 1, but replaces observed time coordinates with **hierarchical latent time coordinates**. This approach borrows strength across cells while allowing individual temporal variation.

The probabilistic graphical model now includes hierarchical priors for latent dimensionless time coordinates, with time parameters becoming random variables rather than observed quantities.

```{python}
#| label: fig-latent-time-pgm
#| code-fold: true
#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates.

# hyperparameters for gene-specific parameters
pgm.add_node("mu_init", r"$\mu_{0}$", 0.5, 6, fixed=True)
pgm.add_node("sigma_init", r"$\sigma_{0}^2$", 1.5, 6, fixed=True)
pgm.add_node("mu_gamma", r"$\mu_{\gamma^*}$", 2.5, 6, fixed=True)
pgm.add_node("sigma_gamma", r"$\sigma_{\gamma^*}^2$", 3.5, 6, fixed=True)
pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)

# hyperparameters for hierarchical time structure (moved further left)
pgm.add_node("mu_T", r"$\mu_{T}$", 5.5, 6, fixed=True)
pgm.add_node("sigma_T", r"$\sigma_{T}^2$", 6, 6, fixed=True)
pgm.add_node("mu_t_loc", r"$\mu_{t_{loc}}$", 6.5, 6, fixed=True)
pgm.add_node("sigma_t_loc", r"$\sigma_{t_{loc}}^2$", 7, 6, fixed=True)
pgm.add_node("mu_t_scale", r"$\mu_{t_{scl}}$", 7.5, 6, fixed=True)
pgm.add_node("sigma_t_scale", r"$\sigma_{t_{scl}}^2$", 8, 6, fixed=True)

# hyperparameters for lambda (entering from right edge of cell plate)
pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 6.5, 3.5, fixed=True)
pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 6.5, 2.5, fixed=True)

# hierarchical time parameters (moved further left)
pgm.add_node("T_max_star", r"$T_{M}^*$", 5.5, 5)
pgm.add_node("t_loc", r"$t_{loc}$", 6.25, 5)
pgm.add_node("t_scale", r"$t_{scl}$", 7, 5)

# latent time coordinates and cell-specific parameters (moved further left, lambda enters from right)
pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)

# latent variables for cell-specific outcomes (centered between rows)
pgm.add_node(
    "u_star_ij",
    r"$u^{\ast}_{ij}$",
    2,
    3.75,
    scale=1.0,
    shape="rectangle",
)
pgm.add_node(
    "s_star_ij",
    r"$s^{\ast}_{ij}$",
    4,
    3.75,
    scale=1.0,
    shape="rectangle",
)

# observed data
pgm.add_node(
    "u_obs_ij",
    r"$u_{ij}$",
    2,
    2.25,
    scale=1.0,
    observed=True,
)
pgm.add_node(
    "s_obs_ij",
    r"$s_{ij}$",
    4,
    2.25,
    scale=1.0,
    observed=True,
)

# edges
edge_params = {"head_length": 0.3, "head_width": 0.25, "lw": 0.7}

# gene-specific parameter edges
pgm.add_edge("mu_init", "u_star_0i", plot_params=edge_params)
pgm.add_edge("sigma_init", "u_star_0i", plot_params=edge_params)
pgm.add_edge("mu_init", "s_star_0i", plot_params=edge_params)
pgm.add_edge("sigma_init", "s_star_0i", plot_params=edge_params)
pgm.add_edge("mu_gamma", "gamma_star_i", plot_params=edge_params)
pgm.add_edge("sigma_gamma", "gamma_star_i", plot_params=edge_params)
pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)

# hierarchical time structure edges
pgm.add_edge("mu_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("sigma_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("mu_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("sigma_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("mu_t_scale", "t_scale", plot_params=edge_params)
pgm.add_edge("sigma_t_scale", "t_scale", plot_params=edge_params)

# time hierarchy edges
pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)

# dynamics edges
pgm.add_edge("u_star_0i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("s_star_0i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("u_star_0i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("gamma_star_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)

# observation edges
pgm.add_edge("u_star_ij", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("s_star_ij", "s_obs_ij", plot_params=edge_params)
pgm.add_edge("U_0i", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("U_0i", "s_obs_ij", plot_params=edge_params)
pgm.add_edge("lambda_j", "u_obs_ij", plot_params=edge_params)
pgm.add_edge("lambda_j", "s_obs_ij", plot_params=edge_params)

# plates
pgm.add_plate(
    [0.5, 1.25, 4.5, 4.25],
    label=r"$i \in \{1, \ldots, G\}$",
    shift=-0.1,
    fontsize=12,
)
pgm.add_plate(
    [1.5, 1.75, 4.5, 2.75],
    label=r"$j \in \{1, \ldots, N\}$",
    shift=-0.1,
    fontsize=12,
)

The model extends the dimensionless approach with **hierarchical latent time coordinates**. The key modification is replacing observed time $t^*_j$ with latent time variables that follow a hierarchical prior structure.

**Priors for dimensionless parameters (unchanged from Study 1):**
\begin{align}
    u^*_{0i}, s^*_{0i} &\sim \text{LogNormal}(\mu_{0}, \sigma_{0}^2) \\
    \gamma^*_i &\sim \text{LogNormal}(\mu_{\gamma^*}, \sigma_{\gamma^*}^2) \\
    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2) \\
    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
\end{align}

**Hierarchical priors for latent time coordinates:**
\begin{align}
    T_{max}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-max-prior} \\
    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \label{eq-t-loc-prior} \\
    t_{scale} &\sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}}) \label{eq-t-scale-prior} \\
    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scale}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
    t^*_j &= T_{max}^* \times \tilde{t}_j \quad \text{(deterministic transformation)} \label{eq-t-star-transform}
\end{align}

$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ (which can be negative for cells before the process start or greater than 1 for cells after the typical process duration), and $t^*_j$ represents the **actual dimensionless time coordinate** obtained by scaling the relative position by the global time scale $T_{max}^*$.

**Dimensionless analytical solutions (unchanged):**
The same analytical solutions from Study 1 apply, but now using latent $t^*_j$:
\begin{align}
    u^*_{ij} &= 1 + (u^*_{0i} - 1) e^{-t^*_j} \\
    s^*_{ij} &= \begin{cases}
        \frac{1}{\gamma^*_i} + \left(s^*_{0i} - \xi_i - \frac{1}{\gamma^*_i}\right) e^{-\gamma^*_i t^*_j} + \xi_i e^{-t^*_j} & \text{if } \gamma^*_i \neq 1 \\
        1 + (s^*_{0i} - 1) e^{-t^*_j} + (u^*_{0i} - 1) t^*_j e^{-t^*_j} & \text{if } \gamma^*_i = 1
    \end{cases}
\end{align}

where $\xi_i = \frac{u^*_{0i} - 1}{\gamma^*_i - 1}$.

**Observation model (unchanged):**
\begin{align}
    u_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot u^*_{ij}) \\
    s_{ij} &\sim \text{Poisson}(\lambda_j \cdot U_{0i} \cdot s^*_{ij})
\end{align}

### Hierarchical Time Parameter Interpretability

The hierarchical time structure introduces new parameters with clear biological and statistical interpretations:

#### **Global Time Scale Parameters**

**$T_{max}^*$ (Global dimensionless time scale)**

- **Definition**: Global scaling factor for dimensionless time coordinates
- **Units**: Dimensionless (in units of characteristic splicing time)
- **Interpretation**: Characteristic temporal extent of the biological process
- **Biological meaning**: Represents the typical time scale over which the biological process unfolds
- **Model role**: Multiplicative scaling factor in $t^*_j = T_{max}^* \times \tilde{t}_j$
- **Prior**: $T_{max}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
- **Example**: $T_{max}^* = 5$ means the process typically spans ~5 characteristic splicing times

#### **Population-Level Time Parameters**

**$t_{loc}$ (Population relative time location)**

- **Definition**: Population-level location parameter for relative time distribution
- **Units**: Dimensionless (relative time units)
- **Interpretation**: Central tendency of where cells are located in relative time
- **Biological meaning**: Typical relative temporal position of cells in the biological process
- **Prior**: $t_{loc} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}})$ ensures positive location with flexible shape
- **Statistical role**: Allows borrowing strength across cells while permitting individual variation
- **Example**: $t_{loc} = 0.7$ means cells are typically positioned at 70% through the relative process timeline

**$t_{scale}$ (Population relative time scale)**

- **Definition**: Population-level scale parameter for relative time distribution
- **Units**: Dimensionless (relative time units)
- **Interpretation**: How much relative temporal variation exists across cells
- **Biological meaning**: Degree of temporal heterogeneity in the cell population
- **Prior**: $t_{scale} \sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}})$ ensures positive scale with flexible shape
- **Statistical role**: Controls the amount of shrinkage toward the population mean
- **Example**: $t_{scale} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$

#### **Cell-Specific Time Coordinates**

**$\tilde{t}_j$ (Cell-specific relative time)**

- **Definition**: Latent relative time coordinate for cell $j$
- **Units**: Dimensionless (relative time units)
- **Interpretation**: Cell's position relative to the population temporal distribution
- **Biological meaning**: Where this cell sits in the relative temporal progression
- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scale}^2)$ allows negative values (before process) and values >1 (after process)
- **Flexibility**: Can represent cells outside the typical process timeline
- **Example**: $\tilde{t}_j = -0.2$ means this cell is before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion

**$t^*_j$ (Cell-specific dimensionless time)**

- **Definition**: Actual dimensionless time coordinate for cell $j$
- **Units**: Dimensionless (in units of characteristic splicing time)
- **Interpretation**: How many splicing events have elapsed for this specific cell
- **Biological meaning**: Cell's absolute position along the temporal trajectory of the biological process
- **Transformation**: $t^*_j = T_{max}^* \times \tilde{t}_j$ (deterministic scaling)
- **Example**: If $T_{max}^* = 5$ and $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times

### Key Advantages of Gamma/Normal Hierarchical Time Structure

#### **Statistical Benefits**

- **Improved identifiability**: Hierarchical structure provides regularization without redundant parameters
- **Borrowing strength**: Information shared across cells improves parameter estimation
- **Uncertainty quantification**: Full posterior distributions for all time parameters
- **Robustness**: Less sensitive to outlier cells or sparse data
- **No truncation constraints**: Avoids computational complexity of truncated distributions
- **Clear parameter roles**: $T_{max}^*$ controls global scale, $t_{loc}$ and $t_{scale}$ control population distribution

- **Dimensionless time**: Universal interpretation across experimental conditions
- **Population heterogeneity**: Captures natural variation in biological timing
- **Individual trajectories**: Each cell has its own temporal position
- **Flexible temporal range**: Cells can be before ($\tilde{t}_j < 0$) or after ($\tilde{t}_j > 1$) typical process timeline
- **Natural biological variation**: No artificial hard boundaries on temporal positions

#### **Computational Efficiency**

- **Standard distributions**: Gamma and Normal distributions are computationally efficient
- **No truncation**: Avoids complex truncated distribution sampling
- **Analytical solutions**: Fast likelihood evaluation for validation
- **Hierarchical sampling**: Efficient MCMC mixing through population-level parameters
- **Multiplicative structure**: Simple deterministic transformation $t^*_j = T_{max}^* \times \tilde{t}_j$
